---
title: "Final Project"
author: "Fatima Fairfax"
date: "4/2/2022"
output: html_document
---

# IDEA 1: EHR and In-Hospital Death Info

## EHR Basic Adoption Data

From https://www.healthit.gov/sites/default/files/data-brief/2014HospitalAdoptionDataBrief.pdf 

Have actual numbers for 2008 and 2011; then I have threshold data for 2008, 2011, and 2014. I will bring over the actual number data. 

```{r load libraries}
library(tidyverse)
library(modelsummary)
library(fixest)

```


```{r import the EHR data}
ehr_data <- EHR_Adoption

glimpse(ehr_data)
```

### Data Cleaning

1. Make the first row the title row
2. Delete the empty columns
3. Make the titles just the year
4. Convert the characters to percentages
5. Make the data long, each state with a column for 2008 and 2011

```{r First row the title row}
ehr_data <- janitor::row_to_names(ehr_data,1,remove_rows_above = FALSE)
```

```{r delete the empty columns}
ehr_data <- ehr_data |> select(State,'2008 Basic EHR, %','2011 Basic EHR, %')
```

```{r make the titles just the year}
colnames(ehr_data) <- c('State','2008','2011')
```

```{r convert the characters to percentages}

ehr_data <- ehr_data |> mutate(`2008` = as.numeric(sub("%","",`2008`))/100,
                   `2011` = as.numeric(sub("%","",`2011`))/100)

glimpse(ehr_data)
```

*This has a bunch of data I don't want / left over from the original set. I want to filter for values in 'State'*

```{r}
ehr_data <- ehr_data |> filter(State %in% state.name) |> slice(1:50)
```


### Variation Check

```{r}
ehr_data |> mutate(diff = `2011` - `2008`)
```



```{r make the data long}

ehr_data <- gather(ehr_data,"year", "EHR_per",2:3)

```




## Hospital Variables Data

```{r}
hosp_data <- Hosp_Data_Final
```


Select what we need: State, Year, `Total number of discharges: N`, `LOS (length of stay), days (mean)`, `LOS (length of stay), days (median)`, `In-hospital deaths: N`, `In-hospital deaths: %`

```{r}
hosp <- hosp_data |> select(State,Year, `Total number of discharges: N`, `LOS (length of stay), days (mean)`, `LOS (length of stay), days (median)`, `In-hospital deaths: N`, `In-hospital deaths: %`)
```


Select years we need: 2008 and 2011

```{r}
hosp <- hosp |> filter(Year %in% c(2008,2011))
```


## Combine the Data

Bring the two data sets together

```{r}
#change the year in hosp to be characters

hosp <- hosp |> mutate(Year = as.character(Year))
```

```{r}
library(reshape2)
```

```{r}
db <- merge(ehr_data,hosp, by.x = c('State','year'), by.y = c('State','Year'))
```


## Rename variables

For ease of use

```{r}
db <- db |> mutate(LOS_mean = `LOS (length of stay), days (mean)`,
                   LOS_med = `LOS (length of stay), days (median)`,
                   death_n = `In-hospital deaths: N`,
                   death_per = `In-hospital deaths: %`,
                   discharge = `Total number of discharges: N`) |> 
  select(-c(`LOS (length of stay), days (mean)`,`LOS (length of stay), days (median)`,`In-hospital deaths: N`,`In-hospital deaths: %`,`Total number of discharges: N`))
```






# IDEA 2: HINTS Data

```{r}
library(haven)

hints_5_1 <- read_sas('hints5_cycle1_public.sas7bdat')
hints_5_2 <- read_sas('hints5_cycle2_public.sas7bdat')
hints_5_3 <- read_sas('hints5_cycle3_public.sas7bdat')
hints_5_4 <- read_sas('hints5_cycle4_public.sas7bdat')
```

Cycle 1: 2017
Cycle 2: 2018
Cycle 3: 2019
Cycle 4: 2020

## Variables of Interest

*TabletHealthWellnessApps - on your tablet or smartphone, do you have apps related to health?*

  -9 - missing
  1 - yes
  2 - no
  3 - don't knopw
  4 - don't have a tablet or smartphone

*Tablet_MakeDecision - has your tablet helped you make a decision about how to treat an illness or condition?*

  -9 - missing data
  -2 - question error
  -1 - inapplicable
  1 - yes
  2 - no

*WearableDevTrackHealth - in the past 12 months, have you used an electronic wearable device to monitor or track your health or activity?*

*OwnAbilityTakeCareHealth - overall, how confident are you about your ability to take good care of your health?*

  -9 - missing
  -5 - multiple in error
  1 - completely confident
  2 - very confident
  3 - somewhat confident
  4 - a little confident
  5 - not confident at all

*GeneralHealth - in general, would you say your health is: 1 - excellent, 5 - poor*

  -9 - missing
  -5 - multiple races in error
  1 - excellent
  2 - very good
  3 - good
  4 - fair
  5 - poor

*MedConditions_Diabetes - has a doctor or health professional ever told you that you had diabetes or high blood sugar?*

*MedConditions_HighBP*

*MedConditions_HeartCondition*

*MedConditions_LungDisease*

### Demographics of Interest

*IncomeRanges - total combined household annual income*

  -9 - missing data
  1 - 0k - 9,999
  2 - 10k - 14,999
  3 - 15k - 19,999
  4 - 30k - 34,999
  5 - 35k - 49,999
  6 - 50k - 74,999
  7 - 75k - 99,999
  8 - 100k - 199,999
  9 - 200k + 
  

*Race_Cat2 - race*

  -9 - missing data
  11 - white
  12 - black
  14 - ameircan indian / alska native
  16 - multiple races selected
  31 - asian indian
  32 - chinese
  33 - filipino
  34 - japanese
  35 - korean
  36 - vietnamese
  37 - other asian
  52 - guamanian or chamorro
  54 - other pacific islander

#### Race Variables 

-9 = missing; -7 = missing; 1 = selected, 2 = not selected

White

Black

AmerInd - american indian or native american

Asian American
  AsInd - asian indian
  Chinese
  Filipino
  Japanese
  Korean
  Vietnamese
  OthAsian

Pacific Islander - one of these
  Hawaiian
  Guamanian
  Samoan
  OthPascIsl

Hisp_Cat - > 0  
  Mexican
  PuertoRican
  Cuban
  OtherHisp

#### Race Variables - Final Edit

RaceEthn5

  -9 - missing
  -7 - missing
  1 - non-hispanic white
  2 - non-hispanic black
  3 - hispanic
  4 - non-hispanic asian
  5 - non-hispanic other

*Age*

  -9 - missing
  -4 - non-conforming
  Rest is current age in years 
  
  
*Education - what is the highest grade or level of schooling you completed?*

  -9 - missing
  1 - less than 8 years (middle school)
  2 - 8 to 11 years (some HS)
  3 - 12 years or completed HS
  4 - post high school other than college
  5 - some college
  6 - college graduate
  7 - postgraduate
  

*SelfGender*

  -9 - missing
  1 - male
  2 - female



### Other possible variables
ProviderMaintainEMR2 - Do any of your doctors/HCP maintain your medical records in a computerized system?
NotAccessed_ConcernedPrivacy - have not accessed online medical record in the last 12 months because you are concerned about the privacy/security of the website for your medical records?

ConfidentGetHealthInf - how confident are you that you could get advice/info about health/medical topics?

TrustDoctor - trust health/medical info from a doctor?

OwnAbilityTakeCareHealth - overall, how confident are you about your ability to take good care of your health?

QualityCare - overall, how would you rate the quality of health care you received in the past 12 months?

GeneralHealth - in general, would you say your health is: 1 - excellent, 5 - poor









## Bring in Selected Variables (cycle 3) - most recent pre-covid

```{r}
db <- hints_5_3 |>  select(TabletHealthWellnessApps,Tablet_MakeDecision,WearableDevTrackHealth,GeneralHealth,IncomeRanges,RaceEthn5,Age,HealthInsurance,Education,SelfGender,MedConditions_Diabetes,MedConditions_HighBP,MedConditions_HeartCondition,MedConditions_LungDisease)
```

## Wrangling

We will do two model sets - the first set with the main independent variable of having a wellness app at all and the second using the app to make decisions.

1. Only select the independent variables with good data. Each we only want 1 = yes or 2 = no
2. Only select those with the dependent variable - GeneralHealth (responses over 0)
3. Rest of the variables are here for matching

### Treatment Variables

```{r treatment variables editing}
app_health <- db |> select(TabletHealthWellnessApps,GeneralHealth,IncomeRanges,RaceEthn5,Age,HealthInsurance,Education,SelfGender) |> 
  filter(TabletHealthWellnessApps == 1 | TabletHealthWellnessApps == 2) |> #only values of yes and no
  filter(GeneralHealth > 0) |> #only take viable values
  mutate(Apps = if_else(TabletHealthWellnessApps == 1, 1L,0L))

dec_health <- db |> select(Tablet_MakeDecision,GeneralHealth,IncomeRanges,RaceEthn5,Age,HealthInsurance,Education,SelfGender) |> 
  filter(Tablet_MakeDecision == 1 | Tablet_MakeDecision == 2) |> 
  filter(GeneralHealth >0) |> 
  mutate(Decision = if_else(Tablet_MakeDecision == 1,1L,0L))

wear_health <- db |> select(WearableDevTrackHealth,GeneralHealth,IncomeRanges,RaceEthn5,Age,HealthInsurance,Education,SelfGender) |> 
  filter(WearableDevTrackHealth == 1 | WearableDevTrackHealth == 2) |> 
  filter(GeneralHealth >0) |> 
  mutate(Wear = if_else(WearableDevTrackHealth == 1,1L,0L))
```

### Matching Variables

```{r variable editing app health}
app_health <- app_health |> mutate(white = if_else(RaceEthn5 == 1, 1L,0L),
                                   black = if_else(RaceEthn5 == 2, 1L,0L),
                                   hisp = if_else(RaceEthn5 == 3,1L,0L),
                                   asian = if_else(RaceEthn5 == 4,1L,0L),
                                   other_race = if_else(RaceEthn5 == 5,1L,0L)) |> #race variables
  filter(IncomeRanges > 0,
         Age > 0,
         Education > 0,
         SelfGender > 0) |> #filter variables to rid missing
  mutate(female = if_else(SelfGender == 2, 1L,0L)) |> #code the gender variable
  mutate(health = 6 - GeneralHealth) |> #flip the health scale so 5 is excellent and 1 is poor
  select(-c(RaceEthn5, SelfGender,GeneralHealth))
  
```

```{r variable editing decision health}
dec_health <- dec_health |> mutate(white = if_else(RaceEthn5 == 1, 1L,0L),
                                   black = if_else(RaceEthn5 == 2, 1L,0L),
                                   hisp = if_else(RaceEthn5 == 3,1L,0L),
                                   asian = if_else(RaceEthn5 == 4,1L,0L),
                                   other_race = if_else(RaceEthn5 == 5,1L,0L)) |> #race variables
  filter(IncomeRanges > 0,
         Age > 0,
         Education > 0,
         SelfGender > 0) |> #filter variables to rid missing
  mutate(female = if_else(SelfGender == 2, 1L,0L)) |> #code the gender variable
   mutate(health = 6 - GeneralHealth) |> #flip the health scale so 5 is excellent and 1 is poor
  select(-c(RaceEthn5, SelfGender,GeneralHealth))
```

```{r variable editing wear health}
wear_health <- wear_health |> mutate(white = if_else(RaceEthn5 == 1, 1L,0L),
                                   black = if_else(RaceEthn5 == 2, 1L,0L),
                                   hisp = if_else(RaceEthn5 == 3,1L,0L),
                                   asian = if_else(RaceEthn5 == 4,1L,0L),
                                   other_race = if_else(RaceEthn5 == 5,1L,0L)) |> #race variables
  filter(IncomeRanges > 0,
         Age > 0,
         Education > 0,
         SelfGender > 0) |> #filter variables to rid missing
  mutate(female = if_else(SelfGender == 2, 1L,0L)) |> #code the gender variable
   mutate(health = 6 - GeneralHealth) |> #flip the health scale so 5 is excellent and 1 is poor
  select(-c(RaceEthn5, SelfGender,GeneralHealth))
```

