---
title: "TE Chpt 6 & 7 - Casual Diagrams"
author: "Fatima Fairfax"
date: "1/22/2022"
output: html_document
---

# Causal Diagrams - Chapter 6

## Question 1

In a conversation with your friend, you mention a study you read that found that being tall causally makes you more likely to earn above $100,000 per year. Your friend says the study must be wrong, since they know several tall people who make much less than that, and several short people who do earn that much. Does your friend’s reasoning make sense or not, and why?

  My friend does not have the right reasoning, though they are right to be skeptical of the conclusion of the study. The fact that they know some tall people who make less money and short people that make more doesn't have any bearing on the fact that at the aggregate, being tall makes it *more* likely that you make over $100k a year. There are definitely other factors that contribute to the likelihood that any individual person makes over $100k, so her short friends make have other features that positively correlate with money and her tall friends may have other ones that negatively correlate with money. If she were to look across her friends and control for other types of variation (e.g., family wealth, education attainment, occupation) then she might be able to challenge the study I found.


## Question 2

Consider the diagram:

```{r}
library(DiagrammeR)
library(tidyverse)
```

```{r Terry Flip Chart}
flip_chart <- 
  create_graph() |> #initiate graph
  add_n_nodes(
    n = 5,
    type = "path",
    label = c("TerrysMood", "CoinFlip", "TerryInRoom", "Cake", "Money"), #Labels for each node
    node_aes = node_aes(
      shape = "circle",
      height = 1,
      width = 1,
      x = c(0,3,3,6,6), #Just the heights of each node (so it looks like a tree)
      y = c(0,2,-2,2,0),
      fill = c("blue", "blue", "blue", "blue", "blue"))) |> 
  add_edge(
    from = 1,
    to = 3) |> 
  add_edge(
      from = 2,
      to = 4) |> 
  add_edge(
      from = 2,
      to = 5) |>  
  add_edge(
      from = 3,
      to = 5)

render_graph(flip_chart)
```


Which variables have a direct effect on money? -- CoinFlip and TerryInRoom have direct effects on money.

Which variables have an indirect effect on money? -- Terry's Mood has an indirect effect on Money
  

## Question 3

"Does reading Harry Potter as a child make you read more as an adult?"

```{r Potter Chart}
hp_chart <- 
  create_graph() |> #initiate graph
  add_n_nodes(
    n = 5,
    type = "path",
    label = c("AgePotterReleased", "HarryPotterAsKid", "LikesReading", "ReadOtherRowlingBooks", "ReadingAsAdult"), #Labels for each node
    node_aes = node_aes(
      shape = "circle",
      height = 0.5,
      width = 1.5,
      x = c(0,0,3,3,6), #Just the heights of each node (so it looks like a tree)
      y = c(3,0,2,-2,0),
      fill = c("blue", "blue", "grey", "blue", "blue"))) |> 
  add_edge(
    from = 1,
    to = 2) |> 
  add_edge(
      from = 2,
      to = 4) |> 
  add_edge(
      from = 2,
      to = 5) |>  
  add_edge(
      from = 3,
      to = 2,
      edge_aes = edge_aes(
        color = "gray")) |> 
  add_edge(
    from = 4,
    to = 5
  ) |> 
  add_edge(
    from = 3,
    to = 5,
    edge_aes = edge_aes(
    color = "gray"
  ))

render_graph(hp_chart)
```


a.	What direct effects should be included when trying to answer your research question of interest?

The direct effects included in this diagram are "likesreading," "readotherrowlingbooks," and "harrypotteraskid." Another thing that could be included is "liked harry potter" or books of that genre. If a child reads harry potter and does not like the book, it might deter them from reading more books. 

b.	What indirect effects should be included when trying to answer your research question of interest?

The indirect effect in this diagram is "agepotterreleased." This is relevant because Harry Potter may have been released when people are already out of childhood, so this would have an effect on if there is a relationship. 

Another thing to be considered is language. If harry potter was not available in a child's native language during childhood then that would have an effect on this diagram as well. 


c.	What is a likely alternative explanation of why we might see a relationship between reading Harry Potter and reading more as an adult?

Besides generally liking reading, reading more as an adult and reading Harry Potter could be due to engagement in a school. Perhaps schools assign books, including Harry Potter, and formal education also leads to adults being in positions where they have to read as an adult. 


## Question 4

a. Why do we bother to include variables on our diagram if we can't observe them?

There is no way we can see all of the variables that might be part of the causal mechanism between two variables, but even if we can't see or measure it we know that there are many factors that may reasonably have an effect on the variables of interest. We include them in the diagram to account for the fact that we know there may be other influences that we can identify but can't observe.

b. Why might we think that LikesReading is an unobserved or latent variable?

Liking reading would make it more likely that a child would read a lot of things, including something as popular as Harry Potter, and that trait would likely also carry them through life where they read more as an adult. This variable would effect both variables and would overstate the effect of of reading Harry Potter on reading as adult.


## Question 5

Consider the research question: "do government fire-safety ads reduce forest fires?"

a. Draw a causal diagram with:
    (a) ForestFires is caused by FireSafetyAds
    (b) FireSafetyAds and ParkRangers are caused by GovtBudget (the government has to pay for this stuff!)
    (c) GovtBudget is caused by TimeSinceLastFire— a recent fire might get more money sent to the parks department

```{r Fire Chart}
fire_chart <- 
  create_graph() |> #initiate graph
  add_n_nodes(
    n = 5,
    type = "path",
    label = c("TimeSinceLastFire", "GovtBudget", "FireSafetyAds", "ParkRangers", "ForestFires"), #Labels for each node
    node_aes = node_aes(
      shape = "circle",
      height = 1,
      width = 1,
      x = c(0,2,4,4,6), #Just the heights of each node (so it looks like a tree)
      y = c(0,0,2,-2,0),
      fill = c("purple", "purple", "purple", "purple", "purple"))) |> 
  add_edge(
    from = 1,
    to = 2) |> 
  add_edge(
      from = 2,
      to = 3) |> 
  add_edge(
      from = 2,
      to = 4) |>
  add_edge(
    from = 3,
    to = 5)

render_graph(fire_chart)
```



b. Suggest one omitted arrow or variable that should be on the diagram

In methods, we would have signs on these arrows and in that scenario there would be a negative arrow between ParkRangers and ForestFires. 

Another (positive) arrow that might exist is between TimeSinceLastFire and ForestFires. Perhaps there are down periods after fires where the conditions are better or worse from more fires. 

Another variable might be gender reveal parties...


## Question 6

Think of a research question: Does parental income have a positive effect on child's high school GPA?

a. Describe the DGP for this question:

Parents likely contribute whatever resources they have to their child's education. The more resources they have, the more parents can contribute. A child's GPA is not just dependent on inherent ability, but resources make a difference in a child's academic achievement. Parent's with resources live in like places which may have an effect on the overall environment that a child learns in. 


b. The cause of interest is **Parental income** and the outcome is **Child's High School GPA**.

c. What other variables are in play?

  - school quality
  - teacher student ratios
  - school district
  - tutoring opportunities
  - parental education level
  - intelligence

d. Draw a causal diagram depicting the relationship between all of the variables

```{r Income to GPA diagram}
ed_chart <- 
  create_graph() |> 
  add_n_nodes(
    n = 7,
    type = "path",
    label = c("ParentalEducation","ParentalIncome","SchoolDistrict","Tutoring","TeacherRatios","ChildGPA","Intelligence"),
    node_aes = node_aes(
      shape = "circle",
      height = 0.5,
      width = 1.5,
      x = c(0,0,3,3,3,6,3),
      y = c(2,0,0,2,-2,0,-4),
      fill = c("darkgreen","darkgreen","darkgreen","darkgreen","darkgreen","darkgreen","gray"))) |> 
  add_edge(
    from = 1,
    to = 2
  ) |> 
  add_edge(
    from = 2,
    to = 3
  ) |> 
  add_edge(
    from = 2,
    to = 5
  ) |> 
  add_edge(
    from = 3,
    to = 4
  ) |> 
  add_edge(
    from = 3,
    to = 5
  ) |> 
  add_edge(
    from = 4,
    to = 6
  ) |> 
  add_edge(
    from = 5,
    to = 6
  ) |> 
  add_edge(
    from = 7,
    to = 2
  ) |> 
  add_edge(
    from = 7,
    to = 6
  )

render_graph(ed_chart)
```


e. Can there be any unobserved or latent variables? Include them in the diagram.

One unobserved variable might be intelligence. In a meritocracy, intelligence may lead to  a parent being in a position to have a high income and if intelligence can be passed down to children / is an innate trait, that would also be relevant in the relationship between income and GPA. 


## Question 7

Consider the research question: "Does the inclusion of 'free shipping' cause people to buy items from an online store more?"

a. Six variables that should be included in a causal diagram

  - 'free shipping' label
  - purchase of online items
  - disposable income
  - distance from physical store
  - public transportation availability
  - owning a car

b. Is it feasible to collect data on all the variables that you listed in part a? Can the variables be measured easily?

You could collect data on all of these variables through a survey and/or online purchase history. They are all reasonably easy to measure. The presence or absence of 'free shipping' and the subsequent purchase can be taken from the purchase history. Income and car possession could be self reported. The public transportation availability and distance from the physical store could be measured if the respondent provides an address, though these two would likely take the most work. 


## Question 8

There is no single definition for causality. However, causality can be understood as the relationship between the cause and effect, or the change in probability of an outcome as a result of the change in a condition. Causality is important and interesting because it can help us understand the mechanisms through which outcomes occur which can help us understand the world better and interfere at the right places when we want to change outcomes.


## Question 9

Which of the following describes a representation of a data generating process (DGP) including variables in that DGP and the causal relationships between them?

    d. Casusal diagram
    





# Drawing Causal Diagrams - Chapter 7

## Question 1

In a simplified casual diagram about the DGP of viewership for a TV show, which of these is true?

a.	The diagram should include a variable for “number of celebrities in the cast” -- **True**. This would be a relevant variable to include.

b.	The diagram should contain one variable for “show airs in the evening” and another for “show doesn’t air in the evening” -- **False** this would be redundant information.

c.	The diagram should not contain a variable for “show budget” because budgets are often secret and the researcher can’t measure them -- **False** even if this can't be measured, it is still likely a part of the causal mechanism so it's appropriate to include it in the diagram

d.	The diagram should contain the variable “review score in the Jefferson Weekly,” which is the newspaper published by the students at Jefferson High School, with a readership of about 120 people. -- **False** this variable would be unimportant because the effect, if existent at all, would be very weak since it's only relevant for 120 people. 


## Question 2

Draw a causal diagram for the question: "do long shift hours make doctors give lower-quality care?" that incorporates:
  
a.	Long shift hours affect both how tired doctors are, and how much experience they have, both of which affect the quality of care
b.	How long shifts are is often decided by the hospital the doctor works at. There are plenty of other things about a given hospital that also affect the quality of care, like its funding level, how crowded it is, and so on
c.	New policies that reduce shift times may be implemented at the same time (with the timing determined by some unobservable change in policy preferences) as other policies that also attempt to improve the quality of care


```{r Hospital Diagram}

hosp_chart <- 
  create_graph() |> 
  add_n_nodes(
    n = 6,
    type = "path",
    label = c("LongHours","Tiredness","Experience","HospitalConditions","HealthcarePolicies","LowQualityCare"),
    node_aes = node_aes(
      shape = "square",
      height = 0.5,
      width = 1.5,
      x = c(0,3,3,3,3,6),
      y = c(0,3,1,-1,-3,0),
      fill = c("purple","lightblue","lightblue","lightblue","lightblue","purple"))) |> 
  add_edge(
    from = 1,
    to = 2
  ) |> 
  add_edge(
    from = 1,
    to = 3
  ) |> 
  add_edge(
    from = 2,
    to = 6
  ) |> 
  add_edge(
    from = 3,
    to = 6
  ) |> 
  add_edge(
    from = 1,
    to = 6
  ) |> 
  add_edge(
    from = 4,
    to = 1
  ) |> 
  add_edge(
    from = 4,
    to = 6
  ) |> 
  add_edge(
    from = 5,
    to = 1
  ) |> 
  add_edge(
    from = 5,
    to = 6
  )

render_graph(hosp_chart)

```



## Question 3

Consider the question: Does funding level of public schools affect student achievement for students in your country?

a.	What is the treatment and what is the outcome of interest? 

The treatment is **funding for public schools** and the outcome is **student achievement**

b.	Write down a list of relevant variables. 

  - funding allocation within schools
  - number of teachers
  - student resources (like books, laptops etc)
  - number of students
  - outside tutoring
  - SES of the school district
  - tutoring

c.	Which of the variables in your list in part b are causes of both treatment and outcome?

SES of the school district has an effect on both the treatment and the outcome. Funding for schools are usually driven by property tax, which is higher in more expensive neighborhoods. Also, higher SES families will be able to afford things like tutors for their kids to raise their achievement. They also have resources and influence over how schools function that might increase their child's achievement (deserved or otherwise).


d.	Why might we want to pay extra attention to the variables listed in part c?

The variables in c are confounders which means that they effect both the treatment and outcome. If we don't account for these variables we might be overstating the causal relationship between the treatment and the outcome.

e.	Draw a causal diagram of the variables listed in part b.

```{r school 2 chart}
school2_chart <- 
  create_graph() |> 
  add_n_nodes(
    n = 8,
    type = "path",
    label = c("Funding","Allocation","NumTeachers","StudentResources","NumStudents","Tutoring","DistrictSES","StudentAchieve"),
    node_aes = node_aes(
      shape = "square",
      height = 0.5,
      width = 1.5,
      x = c(0,2,4,3,4,4,2,6),
      y = c(0,2,2,0,-2,-4,-3,0),
      fill = c("purple","blue","blue","blue","blue","blue","blue","purple"))) |> 
  add_edge(
    from = 1,
    to = 2
  ) |> 
  add_edge(
    from = 1,
    to = 3
  ) |> 
  add_edge(
    from = 1,
    to = 4
  ) |> 
  add_edge(
    from = 2,
    to = 8
  ) |> 
  add_edge(
    from = 3,
    to = 8
  ) |> 
  add_edge(
    from = 4,
    to = 8
  ) |> 
  add_edge(
    from = 5,
    to = 8
  ) |> 
  add_edge(
    from = 6,
    to = 8
  ) |> 
  add_edge(
    from = 7,
    to = 8
  ) |> 
  add_edge(
    from = 7,
    to = 1
  ) 

render_graph(school2_chart)
```


f.	Simplify the diagram from part e.

To simplify this, I will take out number of students and tutoring as they only impact the outcome, so are less relevant for this causal relationship. I will also remove Number of teachers as it is somewhat redundant with student resources. 

```{r school 2 revised}

school2rev <- 
  create_graph() |> 
  add_n_nodes(
    n = 5,
    type = "path",
    label = c("Funding","Allocation","StudentResources","DistrictSES","StudentAchieve"),
    node_aes = node_aes(
      shape = "square",
      height = 0.5,
      width = 1.5,
      x = c(0,3,3,3,6),
      y = c(0,2,0,-2,0),
      fill = c("purple","blue","blue","blue","purple"))) |> 
  add_edge(
    from = 1,
    to = 2
  ) |> 
  add_edge(
    from = 1,
    to = 3
  ) |> 
  add_edge(
    from = 2,
    to = 5
  ) |>  
  add_edge(
    from = 3,
    to = 5
  ) |> 
  add_edge(
    from = 4,
    to = 5
  ) |> 
  add_edge(
    from = 4,
    to = 1
  ) 

render_graph(school2rev)

```


## Question 4

Describe the kinds of situations that each of the following could be applied to in order to simplify a casual diagram

a. **Unimportance** -- if the presence of a variable has a very small and relatively weak effect on the variables of interest we can remove the variable

b. **Redundancy** -- if multiple variables occupy the same space (i.e., are in the same location of the causal mechanism and have cause and effects to the same variables) they can likely be combined and described together

c. **Mediators** -- if a variable is only in the graph to show the mechanism between two other variables, with no other arrow coming in or out of B then it can likely be removed

d. **Irrelevance** -- if a variable is important to the DGP but irrelevant to the research question of interest then it can probably be taken out of the diagram


## Question 5

How can a causal diagram be modified so as to avoid cyclic relationships?

To get out of a cycle you can add a time component. Time only moves in one direction, which will prevent a cycle. You can also find a source of random variation. If you focus on random variation, the effect can't loop back on itself.


## Question 6

Research Question: Does exposure to information about income inequality make college students more likely to support wealth redistribution policies?


a.	What is the cause variable and what is the outcome variable? 

  **Cause**: exposure to income inequality information
  
  **Outcome**: support of wealth redistribution policies
  
b.	Write down a list of between 5 and 10 relevant variables in the data generating process. 

  - exposure to income inequality info
  - support of wealth redistribution policy
  - self-reported political affiliation
  - self-reported class status
  - household SES / income
  - experience in the workforce
  - parents level of tuition contribution
  - network class heterogeneity

c.	Draw a causal diagram incorporating all the variables from part b.

```{r wealth chart}
wealth_chart <- 
  create_graph() |> 
  add_n_nodes(
    n = 8,
    type = "path",
    label = c("InfoExposure","WorkforceExp","ParentPayTuition","NetworkHeterogeneity","ClassStatus","PoliticalAffil","SES","SupportRedistribution"),
    node_aes = node_aes(
      shape = "square",
      height = 0.5,
      width = 1.5,
      x = c(0,2,2,2,4,6,5,4),
      y = c(0,2,0,-2,3,3,-3,0),
      fill = c("orange2","gray","gray","gray","gray","gray","gray","orange2"))) |> 
  add_edge(
    from = 1,
    to = 8
  ) |> 
  add_edge(
    from = 2,
    to = 1
  ) |> 
  add_edge(
    from = 2,
    to = 8
  ) |> 
  add_edge(
    from = 3,
    to = 1
  ) |> 
  add_edge(
    from = 3,
    to = 8
  ) |> 
  add_edge(
    from = 4,
    to = 1
  ) |> 
  add_edge(
    from = 4,
    to = 8
  ) |> 
  add_edge(
    from = 5,
    to = 8
  ) |> 
  add_edge(
    from = 6,
    to = 8
  ) |> 
  add_edge(
    from = 7,
    to = 8
  ) 

render_graph(wealth_chart)
```


d.	Stop working on this problem for fifteen minutes and do something else. Then come back, look at your causal diagram from part c again, and describe one bad assumption you think it made, or something it left out

A potential bad assumption is that political affiliation and SES do not have a direct effect on exposure to income inequality information. Also, that SES does not have a direct effect on parents tuition contribution. SES also likely has a bearing on network class heterogeneity that is missing as well. It's also missing something about knowledge of or about policies that would be relevant for this causal mechanism.


## Question 7

Consider the diagram below. 

```{r Achievement Motivation Cycle}
stumo_chart <- 
  create_graph() |> 
  add_n_nodes(
    n = 2,
    type = "path",
    label = c("Student Achievement","Motivation"),
    node_aes = node_aes(
      shape = "square",
      height = 0.5,
      width = 1.5,
      x = c(0, 4),
      y = c(0,0),
      fill = c("brown","brown"))) |> 
  add_edge(
    from = 1,
    to = 2
  ) |> 
  add_edge(
    from = 2,
    to = 1
  )

render_graph(stumo_chart)

```


It depicts a cyclical relationship between student achievement and motivation. If students achieve more (i.e., score well on exams), then their motivation goes up, and if their motivation goes up, they achieve more. Change the diagram so that the relationship is not cyclic anymore

To break this cycle, we can employ a time dimension. Motivation can come before student achievement or achievement can come before motivation. 

```{r Achievement Motivation Corrected Diagram}

stumo_2 <- 
  create_graph() |> 
  add_n_nodes(
    n = 4,
    type = "path",
    label = c("StudentAchieve_t","Motivation_t","Motivation_t+1","StudentAcheive_t+1"),
    node_aes = node_aes(
      shape = "square",
      height = 0.5,
      width = 1.5,
      x = c(0,0,4,4),
      y = c(2,-2,-2,2),
      fill = c("brown","blue","blue","brown")
    )
  ) |> 
  add_edge(
    from = 1,
    to = 3
  ) |> 
  add_edge(
    from = 2,
    to = 4
  )

render_graph(stumo_2)
```


